Metabolites reflect variability introduced by mesoscale eddies in the North Pacific Subtropical Gyre

### 2024-03-12 manuscript/manuscript.Rmd
Manuscript as sent to coauthors

---
title: "Metabolites reflect variability introduced by mesoscale eddies in the North Pacific Subtropical Gyre"
author: "William Kumler, Laura T. Carlson, Wei Qin, Rachel A. Lundeen, Benedetto Barone, and Anitra E. Ingalls"
date: "`r Sys.Date()`"
output: 
  word_document:
    reference_docx: template.docx
bibliography:
- Exported Items.bib
- r-packages.bib
---

@@

TBD

@@

High frequency, long-term observations at Station ALOHA in the North Pacific Subtropical Gyre (NPSG) over the past 25 years have revealed temporal and spatial variability in what had prevously been considered a homogenous environment [@Karl1999; @Karl2017; @Karl2021]. A major source of variability comes in the form of mesoscale eddies in which water is entrained into circular surface currents tens to hundreds of kilometers in diameter [@Karl2017; @McGillicuddy2016]. These mesoscale features introduce nutrients into the euphotic zone and alter the community composition [@Sweeney2003; @Barone2019]. In particular, the depth and intensity of the deep chlorophyll maximum (DCM) at the top of the nitricline differs dramatically between cyclonic and anticyclonic eddies. @Barone2022 reported a shift in the microbial community at the DCM with eukaryotic phytoplankton enriched and prokaryotic phytoplankton depleted in cyclonic eddies, and @Hawco2021 and @Dugenne2023 showed that some of this difference was due to alterations in dissolved iron chemistry and nitrogen fixation.

@@

Using data for all molecular features, we found that the metabolome differed between the cyclone and anticyclone samples, with multivariate statistics detecting the largest differences in the DCM and 175 meter samples (Figure 1). At the DCM, approximately 16% of the variation in the metabolome could be explained by SLA alone according to the PERMANOVA (p = 0.003) and SLA was strongly correlated with the first principal component (r = .615, % variance explained by PC1 = 27.2), indicating that this was one of the largest sources of variation in the dataset. Notably, the samples taken from the exact center of the cyclonic eddy were highly distinct and likely drove much of the variance explained. In the 175m samples, 12% of the variance was explained by SLA (p = 0.008) according to the PERMANOVA and SLA was highly correlated with the second PC of the dataset (r = .697, % variance explained by PC2 = 15.4), though the first PC did not seem to have any visible pattern with metadata and appeared to capture largely variation between biological triplicates (Supplemental figure 1). At the surface, SLA trends are less evident with a larger p-value (0.087) and a lower R$^2$ (0.09) (Figure 1).

To characterize this difference, we plotted an "average" metabolite peak area per station at each depth by taking the median value of all the detected mass features after z-score normalization (Figure 1). As expected, the average metabolite collected at the exact center of the cyclonic eddy (Station 12) was much more abundant than in any sample at this same depth in the rest of the dipole samples. Similarly, the 175 meter samples showed the largest metabolite abundances in the anticyclone relative to other stations at this depth. This method also allowed us to see the remarkably sinusoidal response of the average metabolite across the stations. This response correlated negatively with SLA at the DCM (lower SLA = higher average values, $r_{DCM}= -.716$) but positively with SLA at the surface and at depth ($r_{surface} = .211$, $r_{deep} = .830$).

@@

We expected a large part of the difference in metabolomes with respect to SLA to be due to the increased biomass in the cyclone DCM and the parallel increase in the anticyclone deep samples. Within our molecular feature population, a subset are known metabolites that we quantified using standards (Figure 2).

@@

We found that the overall concentration of known molecules was highest at the DCM of the cyclone, in line with our expectations and prior results. The most abundant quantified molecule overall was trimethylamine N-oxide (TMAO), with an average concentration of 1.38 nM and a distinct DCM maximum (Figure 2). This pattern of a subsurface maximum at the DCM was mirrored by a majority of the known molecules (37/53). Most other abundant molecules decreased monotonically with depth, with highest values at the surface (Figure 2).

```{r PC_vs_totnM regression values, eval=FALSE}
no_outlier_frac_pc_in_metab_df <- frac_pc_in_metab_df %>%
  filter(!(station=="11" & tripl=="A" & depth=="Surface")) %>%
  filter(!(station=="6" & tripl=="B" & depth=="Surface")) %>%
  filter(!(station=="7" & tripl=="C" & depth=="Surface")) %>%
  filter(!(station=="11" & tripl=="B" & depth=="DCM"))

@@

Known metabolites contributed between 1.5% and 5% of the carbon to the total PC load with higher values at the DCM than the surface or 175m samples (Figure 2, Supplemental Figure 3). DCM samples contained approximately 25% (0.5 percentage points) more carbon in the known metabolite pool (One-way ANOVA with outliers removed plus Tukey post-hoc, p<0.001, n=30-33) but there was no clear trend with SLA.

@@

One sample, taken from the surface at Station 11 (MS11C215m_A), was highly distinct from all others (including its paired triplicates) with several metabolites measuring two or more orders of magnitude larger in peak area and concentration both before and after normalization to internal standards. Heuristic exploration of potentially related masses identified a total of 14 compounds highly abundant in this one sample alone. The only identified compounds with this trend were taurine and acetyltaurine, though three others with *m/z* 146.1176, 148.0968, and 206.1388 were discovered to be likely isomers of known compounds (acetylcholine/(3-carboxypropyl)trimethylammonium, hydroxyisoleucine, and dexpanthenol, respectively). One feature (*m/z* = 184.0637, RT = 7.0 minutes) was identified as choline sulfate using a second run of standards, while an additional mass feature at 168.0689 and RT = 8.5 was putatively flagged as taurine betaine. This mass was also one of the most highly enriched in the MS11C215m_A sample after taurine itself, with a peak area nearly 30x that of the next largest sample. No clear explanation emerged explaining why some compounds were hyperabundant in this surface sample, though one possible explanation was the *Trichodesmium* bloom occurring at the same time [@Dugenne2023].

@@

*Table 1: Detected mass features with peak areas in sample MS11C215m_A much larger than the sample with the next-largest peak area. RT = retention time in minutes. Taurine betaine is followed by a ? to denote that this identification is putative. Feature ID is a disambiguating key produced by XCMS processing.*

@@

We explored this *Trichodesmium* hypothesis further by analyzing additional samples taken from the bloom to see whether compounds abundant in hand-picked *Trichodesmium* colonies were also abundant in samples taken directly from the bloom itself. After searching for the 14 mass features above in the environmental sample, 7 had good-quality peaks when viewed in the bloom samples as extracted ion chromatograms, though only 4 of these matched closely in retention time. We also performed a search of the untargeted data for homoserine betaine, a known *T. erythraeum* (strain IMS101) osmolyte [@Pade2016]. Although this compound was not in our standard library due to a lack of authentic standards, we identified three mass features in the 162.1125 *m/z* window with retention times of 7.1, 8.4, and 9.9 minutes. The feature at RT 9.9 matched our carnitine standard's retention time, leaving the two earlier peaks with putative identifications of threonine betaine at 7.1 and homoserine betaine at 8.4 given the elution order of threonine before homoserine in our authentic standards. This order is the reverse of that in @Pade2016, which showed homoserine betaine eluting prior to carnitine as expected given our use of a polar HILIC column and their use of a nonpolar Hypercarb column. They did not report an additional peak detected at this mass, leaving our putative identification of the threonine betaine as low-confidence. None of the three 162.1125 features showed the same degree of enrichment in MS11C215m_A as taurine or the other 13 compounds detected, though MS11C215m_A was the largest, second-largest, and third-largest peak for carnitine, threonine betaine, and homoserine betaine, respectively.

@@

In addition to absolute shifts in metabolite concentration due to biomass differences, we also explored compositional shifts in the metabolomes.

@@

To provide context for the sea level anomaly responses detected above, we additionally performed high-resolution depth sampling around the DCM at the center of both eddy poles. Here, we found that the ~40 meter difference in depth and the transition between the eddies explained similar amounts of variation in the data (R$^2_{depth}=0.253$, R$^2_{SLA}=0.238$) and were both highly significant factors (permutational p-values << 0.001). Additionally, both sample depth and SLA were well-correlated with the first principal component of the metabolite matrix (r$_{depth}=.766$, r$_{SLA}=.684$, fraction of variance explained by PC$_1$ = 30.8%) (Figure 3). A 2D NMDS plot of the samples showed good separation between the eddies, with cyclonic samples largely on the left side of Figure 3 while anticyclonic samples fell almost entirely to the right. Notably, the cyclonic samples show a much larger spread than the anticyclonic ones, indicating larger sample variability at the surface than the DCM. A depth gradient is visible along MDS 2, with the deepest samples generally at the top right of the plot and the shallowest ones closer to the bottom and left.

```{r figure MC NMDS plot}
MC_nmds_gp <- multivar_analysis_MC$nmds_output %>%
  pluck(1, "points") %>% 
  as.data.frame() %>%
  rownames_to_column("filename") %>%
  left_join(filled_file_metadata %>% distinct(filename, sla_class, depth, tripl)) %>%
  mutate(depth=factor(depth, levels=c("DCMless20m", "DCMless10m", "DCM", 
                                      "DCMplus10m", "DCMplus20m"),
                      labels=c("-20m", "-10m", "DCM", "+10m", "+20m"))) %>%
  ggplot(aes(label=tripl)) +
  geom_vline(xintercept = 0) +
  geom_hline(yintercept = 0) +
  geom_point(aes(x=MDS1, y=MDS2, color=sla_class), size=4, pch=19) +
  scale_color_manual(
    values=c(scales::muted("blue"), scales::muted("red")),
    breaks=c("Cyclone", "Anticyclone"),
  ) +
  annotate("label", x=-200, y=100, label="Cyclone", color=scales::muted("blue"),
           label.r = unit(0, "in"), size=6) +
  annotate("label", x=100, y=120, label="Anticyclone", color=scales::muted("red"),
           label.r = unit(0, "in"), size=6) +
  annotate("label", x=-Inf, y=Inf, label=multivar_analysis_MC$nmds_stress,
           hjust=0, vjust=1, label.r = unit(0, "in")) +
  geom_point(aes(x=MDS1, y=MDS2, fill=depth, alpha=depth), size=4, pch=21, color="black") +
  scale_fill_manual(
    values = c("grey90", "grey80", "grey60", "grey40", "grey20"),
    breaks = c("-20m", "-10m", "DCM", "+10m", "+20m")
  ) +
  scale_alpha_manual(
    values = c(0.4, 0.2, 0, 0.2, 0.4),
    breaks = c("-20m", "-10m", "DCM", "+10m", "+20m")
  ) +
  coord_fixed() +
  guides(alpha="none", color="none", fill=guide_legend(
    title = "Depth", title.position = "top", title.hjust = 0.5, direction = "horizontal"
  )) +
  theme_bw() +
  theme(legend.position = "bottom")
ggsave("MC_nmds_gp.png", plot = MC_nmds_gp, device = "png", 
       path = "figures", width = 6.875, height = 4.5, units = "in", dpi = 300)
```

@@

kclust_df <- kclust %>%
  pluck("cluster") %>%
  data.frame(cluster=.) %>%
  rownames_to_column("feature") %>%
  left_join(untarg_peaks, multiple = "all", by="feature") %>%
  left_join(filled_file_metadata, by="filename") %>%
  filter(ms_run=="MC") %>%
  arrange(desc(abs_depth)) %>%
  mutate(depth=fct_inorder(depth)) %>%
  group_by(feature) %>%
  mutate(norm_area=scale(norm_area)[,1]) %>%
  group_by(cluster, depth, sla_class) %>%
  summarise(mean_rank=mean(norm_area), iqr_value=sd(norm_area), .groups = "drop")
kclust_gp <- kclust_df %>%
  left_join(data.frame(size=kclust$size) %>% mutate(cluster=row_number()), by="cluster") %>%
  mutate(cluster=paste0("Cluster #", cluster, ": n=", size)) %>%
  mutate(depth=factor(depth, labels=rev(c("-20m", "-10m", "DCM", "+10m", "+20m")),
                      levels=rev(c("DCMless20m", "DCMless10m", "DCM", "DCMplus10m", "DCMplus20m")))) %>%
  ggplot(aes(y=depth, color=sla_class)) +
  geom_vline(xintercept = 0) +
  # geom_ribbon(aes(xmin=mean_rank-iqr_value, xmax=mean_rank+iqr_value, group=sla_class),
  #             linewidth=1, fill="#FFFFFF22") +
  geom_path(aes(x=mean_rank, group=sla_class), linewidth=1) +
  geom_point(aes(x=mean_rank), size=2) +
  geom_label(data = data.frame(cluster="Cluster #1: n=73", x=c(0, 0.9), depth=c(5.5, 1),
                               sla_class=c("Anticyclone", "Cyclone")),
             aes(x=x, label=sla_class)) +
  facet_wrap(~cluster, nrow = 1) +
  scale_color_manual(values = c(scales::muted("blue"), scales::muted("red"), "grey50"), 
                     breaks = c("Cyclone", "Anticyclone", "Neither"),
                     guide="none") +
  labs(x="Mean z-scaled metabolite area", y="Sampling depth", color=NULL) +
  coord_cartesian(ylim = c(1, 5.5)) +
  theme_bw() +
  theme(plot.background = element_rect(fill = "white"))
```

@@

fold_change_df <- MC_simpledata %>%
  group_by(feature, station) %>%
  summarise(mean_area=mean(norm_area, na.rm=TRUE)) %>%
  pivot_wider(names_from=station, values_from=mean_area) %>%
  mutate(fold_change=L1/L2) %>%
  select(feature, fold_change)

```{r eddy center volcano plot}
volcanoplot_gp <- pairwise_tests %>%
  left_join(fold_change_df, by = join_by(feature)) %>%
  left_join(rownames_to_column(data.frame(cluster=kclust$cluster), "feature"), 
            by = join_by(feature)) %>%
  left_join(untarg_feats %>% select(feature, compound_name), by = join_by(feature)) %>%
  mutate(known=ifelse(is.na(compound_name), "Unknown", "Known")) %>%
  mutate(compound_name=ifelse(is.na(compound_name), feature, compound_name)) %>%
  ggplot(aes(x=log2(fold_change), y=-log10(p.value), color=factor(cluster), 
             shape=known, label=compound_name)) +
  geom_vline(xintercept = 0) +
  geom_hline(yintercept = 0) +
  geom_hline(yintercept = 1.3, lty=2) +
  geom_point(size=2, stroke=1) +
  scale_y_continuous(breaks = 0:6, labels = paste0("10<sup>-", 0:6, "</sup>")) +
  scale_shape_manual(values=c(4, 16), breaks = c("Unknown", "Known")) +
  scale_color_manual(
    values = c("#414487FF", "#7AD151FF", "#22A884FF", "#2A788EFF"),
    breaks = 1:4
  ) +
  theme_bw() +
  labs(x="Log-2 fold-change cyclone enrichment", y="Mann-Whitney p-value\n(FDR-corrected)",
       shape="Metabolite\nstatus", color="k-means\ncluster") +
  guides(color=guide_legend(order = 2), shape=guide_legend(order=1)) +
  theme(axis.text.y = ggtext::element_markdown(),
        plot.background = element_rect(fill = "white"))
```

```{r kclust volcano plot combine}
kclust_volcano_gp <- egg::ggarrange(
  kclust_gp, volcanoplot_gp + theme(legend.position = "none"), 
  ncol=1, heights = c(2, 1.5)) %>%
  cowplot::plot_grid(
    cowplot::get_legend(volcanoplot_gp), rel_widths = c(0.85, 0.15)
  )
ggsave("kclust_volcano_gp.png", plot = kclust_volcano_gp, device = "png", 
       path = "figures", width = 6.875, height = 4, units = "in", dpi = 300,
       bg="white")
```

@@

*Figure 4: Distribution of metabolites in the high-resolution depth samples from the centers of each eddy. The upper row of plots shows k-means clusters where points denote the average z-scored peak area for both known and unknown metabolites across the samples and are colored by the eddy from which they were taken. Clusters have been ordered by number of metabolites in each group and the total is denoted in the panel titles. Both depth trends (mostly a net decrease in metabolites with depth) and eddy effects (cyclonic enrichment in clusters 1 and 2, anticyclone enrichment in cluster 4) are observable. The lower plot shows the individual known and unknown metabolites where points correspond to the FDR-corrected p-value estimated by the nonparametric Mann-Whitney U test and the log$_2$ fold-change calculated with the average peak area in the cyclone divided by the average peak area in the anticyclone. Colors have been assigned using the k-means clusters and shapes have been assigned based on the status of the mass feature as either a known metabolite that was run with an authentic standard or an unidentified metabolite.*

@@

Many of the known metabolites enriched in the cyclone are expected given the overall higher biomass, particularly as many have functions as osmolytes in the cell. Metabolites enriched in excess of biomass were taurine and isethionic acid, the reduced sulfur osmolytes dimethylsulfoniopropionate (DMSP) and dimethylsulfonioacetate (DMS-Ac), and the isomers homarine and trigonelline. A few metabolites more abundant in the anticyclone were also given putative identifications based on RT and *m/z* matching with internal standards later added. Of these, the putative arsenobetaine was the most significantly different among these with a peak area at the anticyclone DCM nearly quadruple that of the cyclone. Of note, the 173.09211 *m/z* mass feature noted above was among the most significantly different between the two eddies, along with two isomers at an *m/z* of 170.11756 and retention times around 8-9 minutes that increased by a factor of 1.5 in the anticyclone (putative formula C$_7$H$_{14}$N$_4$O).

@@

The North Pacific Subtropical Gyre shows significant variation in its particulate composition due to mesoscale eddies. We sampled the metabolome across two sets of eddy dipoles to explore the effect of sea level anomaly (SLA).

>>>>

---
title: "Metabolites reflect variability introduced by mesoscale eddies in the North Pacific Subtropical Gyre"
author: "William Kumler, Wei Qin, Rachel A. Lundeen, Benedetto Barone, Laura T. Carlson, and Anitra E. Ingalls"
date: "`r Sys.Date()`"
output: 
  word_document:
    reference_docx: template.docx
bibliography:
- Exported Items.bib
- r-packages.bib
csl: limnology-and-oceanography.csl
---

@@

Mesoscale eddies significantly alter open ocean environments such as the North Pacific Subtropical Gyre. Previous studies have explored eddy effects on biogeochemical measurements and community composition but not on the molecular composition of particulate organic matter. This study reports the exact concentration of 67 metabolites and relative abundances for 640 molecular features to understand how mesoscale eddies impact the NPSG's metabolome during two cruises in 2017 and 2018. We find that many metabolites track biomass trends, but biosignals like isethionic acid, homarine, and trigonelline linked to picoeukaryotes were enriched at the deep chlorophyll maximum of the cyclonic features, while degradation products such as arsenobetaine were enriched in anticyclones. Additionally, we identify taurine betaine (N,N,N-trimethyltaurine) in open ocean samples for the first time. By analyzing depth variability (accounting for 20-40% of metabolomic variability across ~150 meters) and isopycnal disruption (explaining 10-20% of variability across a sea level anomaly range of 40 centimeters), this analysis constrains the importance of mesoscale eddies in modeling the chemical composition of particulate matter in the largest biomes on the planet.

@@

High frequency, long-term observations at Station ALOHA in the North Pacific Subtropical Gyre (NPSG) over the past 25 years have revealed temporal and spatial variability in what had prevously been considered a homogenous environment [@Karl1999; @Karl2017; @Karl2021]. A major source of variability comes in the form of mesoscale eddies in which water is entrained into circular surface currents tens to hundreds of kilometers in diameter [@Karl2017; @McGillicuddy2016; @Rii2022]. These mesoscale features introduce nutrients into the euphotic zone and alter the community composition [@Sweeney2003; @Barone2019]. In particular, the depth and intensity of the deep chlorophyll maximum (DCM) at the top of the nitricline differs dramatically between cyclonic and anticyclonic eddies. @Barone2022 reported a shift in the microbial community at the DCM with eukaryotic phytoplankton enriched and prokaryotic phytoplankton depleted in cyclonic eddies, and @Hawco2021 and @Dugenne2023 showed that some of this difference was due to alterations in dissolved iron chemistry and nitrogen fixation.

@@

We found that the metabolome clearly differed between the cyclonic and anticyclonic samples when all molecular features were analyzed. Multivariate statistics detected the largest differences in particulate matter composition at the DCM and 175 meter samples (Figure 1). At the DCM, approximately 16% of the variation in the metabolome could be explained by sea level anomaly (SLA) alone (PERMANOVA, p = 0.003) and SLA was strongly correlated with the first principal component (r = .615, % variance explained by PC1 = 27.2), indicating that this was one of the largest sources of variation in the dataset. Notably, the samples taken from the exact center of the cyclonic eddy were highly distinct and likely drove much of the variance explained. In the 175m samples, 12% of the variance was explained by SLA (PERMANOVA, p = 0.008) and SLA was highly correlated with the second PC of the dataset (r = .697, % variance explained by PC2 = 15.4), though the first PC did not seem to have any visible pattern with metadata and appeared to largely capture variation between biological triplicates (Supplemental figure 1). At the surface, SLA trends are less evident with a larger p-value (PERMANOVA, p=0.087) and a lower R$^2$ (0.09) (Figure 1).

To characterize the difference between eddies in a lower-dimensional space, we plotted an "average" metabolite peak area per station at each depth by taking the median value of each of the detected mass features after z-score normalization (Figure 1). As expected, the average metabolite collected at the exact center of the cyclonic eddy (Station 12) was more abundant than in other samples at this same depth. Similarly, the 175 meter samples showed the largest metabolite abundances in the anticyclone relative to other stations at this depth. This method also allowed us to see the remarkably sinusoidal response of the average metabolite across the stations. This response correlated negatively with SLA at the DCM (lower SLA = higher average values, $r_{DCM}= -.716$) but positively with SLA at the surface and at depth ($r_{surface} = .211$, $r_{deep} = .830$).

@@

We expected a large part of the SLA trends to be due to the elevated biomass in the cyclone DCM relative to the anticyclone and the inverse pattern in the 175 meter samples. While we were unable to quantify a majority of the molecular features in the dataset for direct comparison to biomass metrics, a subset of those were known metabolites that we quantified using standards (Figure 2).

@@

We found that the overall concentration of known molecules was highest at the DCM of the cyclone. The most abundant quantified molecule in the eddy transect samples was trimethylamine N-oxide (TMAO), with an average concentration of 1.38 nM and a distinct DCM maximum (Figure 2). This pattern of a subsurface maximum at the DCM was mirrored by a majority of the known molecules (37/53). Most other abundant molecules decreased monotonically with depth, with highest values at the surface.

```{r PC_vs_totnM regression values}
no_outlier_frac_pc_in_metab_df <- frac_pc_in_metab_df %>%
  filter(!(station=="11" & tripl=="A" & depth=="Surface")) %>%
  filter(!(station=="6" & tripl=="B" & depth=="Surface")) %>%
  filter(!(station=="7" & tripl=="C" & depth=="Surface")) %>%
  filter(!(station=="11" & tripl=="B" & depth=="DCM"))

@@

Known metabolites contributed between 1.5% and 5% of the carbon to the total PC load with higher values at the DCM than the surface or 175m samples (Figure 2, Supplemental Figure 3). DCM samples contained approximately 25% more carbon in the known metabolite pool (One-way ANOVA with outliers removed plus Tukey post-hoc, p<0.001, n=30-33) but there was no clear trend with SLA.

@@

One sample, taken from the surface at Station 11 (MS11C215m_A), was highly distinct from all others (including its paired triplicates) with several metabolites measuring two or more orders of magnitude larger in peak area and concentration both before and after normalization to internal standards. Heuristic exploration of potentially related masses identified a total of 14 compounds highly abundant in this one sample alone. The only identified compounds with this trend were taurine and acetyltaurine, though three others with *m/z* 146.1176, 148.0968, and 206.1388 were discovered to be likely isomers of known compounds (acetylcholine/(3-carboxypropyl)trimethylammonium, hydroxyisoleucine, and dexpanthenol, respectively). One feature (*m/z* = 184.0637, RT = 7.0 minutes) was identified as choline sulfate using a second run of standards, while an additional mass feature at 168.0689 and RT = 8.5 was putatively flagged as taurine betaine. This mass was also one of the most highly enriched in the MS11C215m_A sample after taurine itself, with a peak area nearly 30x that of the next largest sample. No clear explanation emerged for why some compounds were hyperabundant in this surface sample, though one possible explanation was the *Trichodesmium* bloom that was also sampled at this time [@Dugenne2023].

@@

*Table 1: Detected mass features that were enriched in the 15m sample from Station 11 (MS11C215m_A). RT = retention time in minutes. Taurine betaine is followed by a ? to denote that this identification is putative. Feature ID is a disambiguating key produced by XCMS processing.*

@@

We explored the possibility that a nearby *Trichodesmium* bloom was responsible for the unusual composition of this one sample by analyzing additional samples taken directly from the bloom to see whether compounds abundant in hand-picked *Trichodesmium* colonies were also abundant the Station 11 samples. After searching for the 14 mass features above in the environmental sample, 7 had good-quality peaks when viewed in the bloom samples as extracted ion chromatograms, though only 4 of these matched closely in retention time. We also performed a search of the untargeted data for homoserine betaine, a known *T. erythraeum* (strain IMS101) osmolyte [@Pade2016]. Although this compound was not in our standard library due to a lack of authentic standards, we identified three mass features in the 162.1125 *m/z* window with retention times of 7.1, 8.4, and 9.9 minutes. The feature at RT 9.9 matched our carnitine standard's retention time, leaving the two earlier peaks with putative identifications of threonine betaine at 7.1 and homoserine betaine at 8.4 given the elution order of threonine before homoserine in our authentic standards. This order is the reverse of that in @Pade2016, which showed homoserine betaine eluting prior to carnitine. This elution order is expected given our use of a normal phse HILIC column and their use of a reversed phase Hypercarb column. They did not report an additional peak detected at this mass, leaving our putative identification of the threonine betaine as low-confidence. None of the three 162.1125 features showed the same degree of enrichment in MS11C215m_A as taurine or the other 13 compounds detected, though MS11C215m_A was the largest, second-largest, and third-largest peak for carnitine, threonine betaine, and homoserine betaine, respectively.

@@

In addition to absolute shifts in metabolite concentration due to biomass differences, we also explored compositional shifts in the metabolomes by assessing the fraction of the total metabolite pool contributed by each compound across the eddy transect.

@@

To provide context for the sea level anomaly responses detected above, we additionally performed high-resolution depth sampling around the DCM at the center of both eddy poles. Here, we found that the ~40 meter depth range and the transition between the eddies explained similar amounts of variation in the data (R$^2_{depth}=0.253$, R$^2_{SLA}=0.238$) and were both highly significant factors (permutational p-values << 0.001). Additionally, both sample depth and SLA were correlated with the first principal component of the metabolite matrix (r$_{depth}=.766$, r$_{SLA}=.684$, fraction of variance explained by PC$_1$ = 30.8%) (Figure 3). A 2D NMDS plot of the samples showed clear separation between the eddies, with cyclonic samples largely on the left side of Figure 3 while anticyclonic samples fell almost entirely to the right. Notably, the cyclonic samples show a much larger spread than the anticyclonic ones, indicating larger sample variability at the surface than the DCM. A depth gradient is visible along MDS 2, with the deepest samples generally at the top right of the plot and the shallowest ones closer to the bottom and left.

```{r figure MC NMDS plot}
MC_nmds_gp <- multivar_analysis_MC$nmds_output %>%
  pluck(1, "points") %>% 
  as.data.frame() %>%
  rownames_to_column("filename") %>%
  left_join(filled_file_metadata %>% distinct(filename, sla_class, depth, tripl)) %>%
  mutate(depth=factor(depth, levels=c("DCMless20m", "DCMless10m", "DCM", 
                                      "DCMplus10m", "DCMplus20m"),
                      labels=c("-20m", "-10m", "DCM", "+10m", "+20m"))) %>%
  ggplot(aes(label=tripl)) +
  geom_vline(xintercept = 0) +
  geom_hline(yintercept = 0) +
  geom_point(aes(x=MDS1, y=MDS2, color=sla_class), size=5, pch=19) +
  scale_color_manual(
    values=c(scales::muted("blue"), scales::muted("red")),
    breaks=c("Cyclone", "Anticyclone"),
  ) +
  annotate("label", x=-200, y=100, label="Cyclone", color=scales::muted("blue"),
           label.r = unit(0, "in"), size=6) +
  annotate("label", x=100, y=120, label="Anticyclone", color=scales::muted("red"),
           label.r = unit(0, "in"), size=6) +
  annotate("label", x=-Inf, y=Inf, label=multivar_analysis_MC$nmds_stress,
           hjust=0, vjust=1, label.r = unit(0, "in")) +
  geom_point(aes(x=MDS1, y=MDS2, fill=depth, alpha=depth), size=5, pch=21, color="black") +
  scale_fill_manual(
    values = c("white", "white", "white", "black", "black"),
    breaks = c("-20m", "-10m", "DCM", "+10m", "+20m")
  ) +
  scale_alpha_manual(
    values = c(0.7, 0.4, 0.2, 0.1, 0.4),
    breaks = c("-20m", "-10m", "DCM", "+10m", "+20m")
  ) +
  coord_fixed() +
  guides(alpha="none", color="none", fill=guide_legend(
    title = "Depth", title.position = "top", title.hjust = 0.5, direction = "horizontal",
    override.aes = list(fill=c("grey90", "grey80", "grey60", "grey40", "grey20"))
  )) +
  theme_bw() +
  theme(legend.position = "bottom")
MC_nmds_gp
ggsave("MC_nmds_gp.png", plot = MC_nmds_gp, device = "png", 
       path = "figures", width = 6.875, height = 4.5, units = "in", dpi = 300)
```

@@

kclust_df <- kclust %>%
  pluck("cluster") %>%
  data.frame(cluster=.) %>%
  rownames_to_column("feature") %>%
  left_join(untarg_peaks, multiple = "all", by="feature") %>%
  left_join(filled_file_metadata, by="filename") %>%
  filter(ms_run=="MC") %>%
  arrange(desc(abs_depth)) %>%
  mutate(depth=fct_inorder(depth)) %>%
  group_by(feature) %>%
  mutate(norm_area=scale(norm_area)[,1]) %>%
  group_by(cluster, depth, sla_class) %>%
  summarise(mean_rank=mean(norm_area), iqr_value=sd(norm_area), .groups = "drop")
library(ggh4x)
strip_theme <- strip_themed(background_x = elem_list_rect(fill= c(
  "#414487FF", "#7AD151FF", "#22A884FF", "#2A788EFF")))
kclust_gp <- kclust_df %>%
  left_join(data.frame(size=kclust$size) %>% mutate(cluster=row_number()), by="cluster") %>%
  mutate(clusterlab=paste0("Cluster #", cluster, ": n=", size)) %>%
  mutate(depth=factor(depth, labels=rev(c("-20m", "-10m", "DCM", "+10m", "+20m")),
                      levels=rev(c("DCMless20m", "DCMless10m", "DCM", "DCMplus10m", "DCMplus20m")))) %>%
  ggplot(aes(y=depth, color=sla_class)) +
  geom_vline(xintercept = 0) +
  geom_path(aes(x=mean_rank, group=sla_class), linewidth=1) +
  geom_point(aes(x=mean_rank), size=3) +
  ggh4x::facet_wrap2(~clusterlab, nrow = 1, strip = strip_theme) +
  scale_color_manual(values = c(scales::muted("blue"), scales::muted("red"), "grey50"), 
                     breaks = c("Cyclone", "Anticyclone", "Neither"),
                     guide="none") +
  labs(x="Mean z-scaled metabolite area", y="Sampling depth", color=NULL) +
  coord_cartesian(ylim = c(1, 5.5)) +
  theme_bw() +
  theme(plot.background = element_rect(fill = "white"),
        strip.background = element_rect(fill = "white"),
        strip.text = element_text(face = "bold", color="white"))
```

@@

fold_change_df <- MC_simpledata %>%
  group_by(feature, station) %>%
  summarise(mean_area=mean(norm_area, na.rm=TRUE)) %>%
  pivot_wider(names_from=station, values_from=mean_area) %>%
  mutate(fold_change=L1/L2) %>%
  select(feature, fold_change)
```

```{r eddy center volcano plot}
volcanoplot_gp <- pairwise_tests %>%
  left_join(fold_change_df, by = join_by(feature)) %>%
  left_join(rownames_to_column(data.frame(cluster=kclust$cluster), "feature"), 
            by = join_by(feature)) %>%
  left_join(untarg_feats %>% select(feature, compound_name), by = join_by(feature)) %>%
  mutate(known=ifelse(is.na(compound_name), "Unknown", "Known")) %>%
  mutate(compound_name=ifelse(is.na(compound_name), feature, compound_name)) %>%
  ggplot(aes(x=log2(fold_change), y=-log10(p.value), color=factor(cluster), 
             shape=known, label=compound_name)) +
  geom_vline(xintercept = 0) +
  geom_hline(yintercept = 0) +
  geom_hline(yintercept = 1.3, lty=2) +
  geom_point(size=2, stroke=1) +
  scale_y_continuous(breaks = 0:6, labels = paste0("10<sup>-", 0:6, "</sup>")) +
  scale_shape_manual(values=c(4, 16), breaks = c("Unknown", "Known")) +
  scale_color_manual(
    values = c("#414487FF", "#7AD151FF", "#22A884FF", "#2A788EFF"),
    breaks = 1:4
  ) +
  theme_bw() +
  labs(x="Log-2 fold-change cyclone enrichment", y="Mann-Whitney p-value\n(FDR-corrected)",
       shape="Metabolite status", color="k-means\ncluster") +
  guides(color="none", shape=guide_legend(order=1)) +
  theme(axis.text.y = ggtext::element_markdown(),
        plot.background = element_rect(fill = "white"),
        legend.position = c(0, 1), legend.justification = c(0, 1), 
        legend.background = element_rect(fill="#FFFFFF88", color=NA),
        legend.spacing.y=unit(0, "in"),
        legend.key.size = unit(0, "cm"))
```

```{r kclust volcano plot combine}
kclust_volcano_gp <- egg::ggarrange(
  kclust_gp, volcanoplot_gp, 
  ncol=1, heights = c(2, 1.5))
ggsave("kclust_volcano_gp.png", plot = kclust_volcano_gp, device = "png", 
       path = "figures", width = 6.875, height = 4, units = "in", dpi = 300,
       bg="white")
```

@@

*Figure 4: Distribution of metabolites in the high-resolution depth samples from the centers of each eddy. The upper row of plots shows k-means clusters where points denote the average z-scored peak area for both known and unknown metabolites across the samples and are colored by the eddy from which they were taken. Clusters have been ordered by number of metabolites in each group and the total is denoted in the panel titles. Both depth trends (mostly a net decrease in metabolites with depth) and eddy effects (cyclonic enrichment in clusters 1 and 2, anticyclone enrichment in cluster 4) are observable. The lower plot shows the individual known and unknown metabolites where points correspond to the FDR-corrected p-value estimated by the nonparametric Mann-Whitney U test and the log$_2$ fold-change calculated with the average peak area in the cyclone divided by the average peak area in the anticyclone. Colors have been assigned using the k-means clusters and shapes have been assigned based on the status of the mass feature as either a known metabolite that was matched to an authentic standard or an unidentified metabolite.*

@@

Many of the known metabolites enriched in the cyclone are expected given the overall higher biomass, particularly as many function as osmolytes in the cell. Metabolites enriched in excess of biomass were taurine and isethionic acid, the reduced sulfur osmolytes dimethylsulfoniopropionate (DMSP) and dimethylsulfonioacetate (DMS-Ac), and the isomers homarine and trigonelline. A few metabolites more abundant in the anticyclone were also given putative identifications based on RT and *m/z* matching with internal standards later added. Of these, the putative arsenobetaine was the most significantly different among these with a peak area at the anticyclone DCM nearly quadruple that of the cyclone. Of note, the 173.09211 *m/z* mass feature noted above was among the most significantly different between the two eddies, along with two isomers at an *m/z* of 170.11756 and retention times around 8-9 minutes that increased by a factor of 1.5 in the anticyclone (putative formula C$_7$H$_{14}$N$_4$O).

@@

We measured the metabolome of samples collected across two sets of eddy dipoles to explore the effect of sea level anomaly (SLA) on metabolite composition and found that mesoscale eddies explain a significant portion of observed variations in particulate metabolomes.
